---
title: "Talking to other systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Talking to other systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(blockr.process)
```

Half of what a real process waits for happens somewhere else. A file lands
on an upload platform. A nightly load finishes. A ticket closes. A lab
returns a result. None of that is a click in your app, and none of it
should have to be.

The interface already exists: the event log **is** the API. Current state
is a fold over an append-only list of events, so causing an event is the
only thing an outside system ever needs to do. This vignette is about the
door it knocks on.

## The inbox

One message per file, dropped into `<store>/inbox/`:

```json
{"instance": "2026Q1", "task": "delivery", "element": "Northgate",
 "field": "status", "value": "done", "actor": "delivery-platform"}
```

Only `task` is required. `instance` defaults to `"latest"`, `field` to
`"status"`, `value` to `"done"`, `actor` to `"external"` — so the smallest
useful message is `{"task": "delivery"}`.

Nothing but a directory is involved: no server, no port, no token store, no
dependency. Every language can write a file, and it works over any shared
or synced mount. What it costs is that **the filesystem is the
authentication** — whoever may write into the inbox may move a process
forward. That is the same trust as "whoever may INSERT into the events
table"; enforce it with directory permissions.

## End to end

```{r}
store <- tempfile()

process <- data.frame(
  task       = c("each_unit", "delivery", "review", "approve"),
  name       = c("for each unit", "Data delivery", "Review", "Approve"),
  role       = c("", "system", "analyst", "management"),
  depends_on = c("", "", "delivery", "each_unit"),
  collection = c("unit", "", "", ""),
  parent     = c("", "each_unit", "each_unit", "")
)

start_instance(process, c("north", "south"), "2026Q1", store)
```

`delivery` is `system` **with no script**: the worker will not run it (no
script) and the task list will not let anyone tick it (role `system`), so
the only thing that can complete it is a message from outside. That is the
convention for "this task waits for the world" — BPMN's receive task.

The platform delivers `north`:

```{r}
write_inbox_message(
  store, task = "delivery", element = "north", instance = "2026Q1",
  actor = "delivery-platform", id = "delivery-2026Q1-north"
)

ingest_inbox(store)
```

```{r}
tab <- instance_view(store, "2026Q1")
tab[, c("task", "element", "status")]
```

`north` may be reviewed; `south` is still blocked, waiting for its own
delivery. The event says who caused it and which message it was:

```{r}
instance_log(store, "2026Q1")[1, c("task", "element", "value", "actor", "ref")]
```

`ref` is what makes an event traceable back **out** of the process: `actor`
says which system spoke, `ref` says which of its messages this was.

In production nobody calls `ingest_inbox()` by hand — `run_worker()` does
it on every tick. A process with no scripts at all can run
`ingest_inbox()` from cron instead; it is an ordinary function.

## Resending is free

Networks time out and senders retry. A delivery counted twice is a wrong
process, so **the file name is the idempotency key**:

```{r}
write_inbox_message(
  store, task = "delivery", element = "north", instance = "2026Q1",
  actor = "delivery-platform", id = "delivery-2026Q1-north"
)

ingest_inbox(store)
sum(instance_events(store, "2026Q1")$task == "delivery")
```

One event, not two. Give a message a stable id that says what it is about
(`delivery-2026Q1-north`), and a resend costs nothing. The duplicate is
kept in `processed/` rather than deleted, because a file someone sent is
evidence.

## Three outcomes

| outcome | what happens | where the file goes |
|---|---|---|
| applied | one event is appended | `inbox/processed/` |
| impossible | nothing is written | `inbox/failed/`, with a `.error` beside it |
| early | nothing happens yet | stays in `inbox/` |

"Impossible" means it can never apply, however often it is retried: no such
task in this instance, an element that is not in the collection, an element
on a task that does not repeat, unreadable JSON. The reason is written next
to the file:

```{r}
write_inbox_message(store, "delivery", element = "westside",
                    instance = "2026Q1", id = "wrong")
ingest_inbox(store, quiet = TRUE)

readLines(file.path(inbox_dir(store), "failed", "wrong.json.error"))
```

"Early" means the message is right but the instance has not been started
yet — a race, not a mistake. The file stays where it is and the next tick
picks it up.

Validation is against **the instance**, not a schema: the definition an
instance was started with is stamped into its log, so the check knows which
tasks exist and which elements are real.

## Sending from somewhere else

Anything that can write a file can send a message. Write to a temporary
name in the same directory and rename, so a half-written file is never
read:

```sh
# shell
cat > /srv/store/inbox/.tmp <<JSON
{"instance":"2026Q1","task":"delivery","element":"north","actor":"platform"}
JSON
mv /srv/store/inbox/.tmp /srv/store/inbox/delivery-2026Q1-north.json
```

```sql
-- Postgres: a trigger writing the same file with COPY ... TO PROGRAM,
-- or (more usual) a NOTIFY that a small listener turns into a message
```

```python
# python
import json, os, tempfile
msg = {"instance": "2026Q1", "task": "delivery", "element": "north",
       "actor": "platform"}
d = "/srv/store/inbox"
fd, tmp = tempfile.mkstemp(dir=d)
os.write(fd, json.dumps(msg).encode()); os.close(fd)
os.rename(tmp, os.path.join(d, "delivery-2026Q1-north.json"))
```

## When the sender cannot reach the filesystem

Two variants, both built on the same message, so the process side does not
change.

**HTTP** — put a [plumber](https://www.rplumber.io) route in front of the
inbox. This is the whole server:

```{r, eval = FALSE}
# plumber.R
library(blockr.process)
store <- Sys.getenv("PROC_STORE")

#* @post /events
function(req, res, task, element = NULL, value = "done",
         field = "status", instance = "latest", actor = "api", id = NULL) {
  if (!identical(req$HTTP_AUTHORIZATION, paste("Bearer", Sys.getenv("PROC_TOKEN")))) {
    res$status <- 401
    return(list(error = "unauthorized"))
  }
  write_inbox_message(store, task, value, field, element, instance, actor, id)
  list(ok = TRUE)
}
```

```sh
curl -X POST https://.../events -H "Authorization: Bearer $TOKEN" \
  -d task=delivery -d element=north -d id=delivery-2026Q1-north
```

The endpoint never writes an event; it writes a message, and the worker
does the rest. Idempotency, validation and the audit trail come along
unchanged.

**Pull** — when nobody may reach in, reach out. A scheduled script that
asks the other system what is true and turns the difference into messages:

```{r, eval = FALSE}
delivered <- DBI::dbGetQuery(con, "
  SELECT unit, uploaded_at FROM deliveries WHERE period = '2026Q1'
")

for (i in seq_len(nrow(delivered))) {
  write_inbox_message(
    store, task = "delivery", element = delivered$unit[i],
    instance = "2026Q1", actor = "upload-db",
    id = paste0("delivery-2026Q1-", delivered$unit[i])   # <- the diff, for free
  )
}
ingest_inbox(store)
```

Note what the id buys: the script does not need to remember what it already
sent, and it does not need to query for "new" rows. Sending everything
every time is correct, because the second send of a message is ignored.
That is usually simpler than a watermark, and it is self-healing.

## Assignments, not just completions

A message sets any field, so a roster system can assign work:

```{r}
write_inbox_message(store, "review", field = "assignee", value = "ana",
                    element = "north", instance = "2026Q1",
                    actor = "hr-system", id = "assign-north")
ingest_inbox(store, quiet = TRUE)

tab <- instance_view(store, "2026Q1")
tab[tab$task == "review", c("element", "assignee", "status")]
```

## What is not here yet

Everything above is inbound. A process that wants to **tell** another
system something — a webhook when an instance completes, a row written back
when a delivery is approved — has no support yet; you would write it as a
script task, which is not wrong but is not a first-class notion either.
It is the most likely next addition.
